# Input / output with *`readr`*

`readr` provides functions for reading in and writing out files.
<br>
Those functions :

* are **up to 10 times faster** that the base R equivalent.
* automatically uncompress files with **.gz**, **.bz2**, **.xz** or **.zip** extensions.
* automatically download files starting with **http://**, **https://**, **ftp://**, or **ftps://**.

## Read

`read_delim`, `read_csv`, `read_tsv`: read a delimited file into a `tibble`:

```{r, echo=FALSE, eval=TRUE}
knitr::kable(
  data.frame(name=c("read_delim", "read_csv", "read_tsv"), 
             delim=c("*needs to be set*", "`,`: comma-separated", "`\t`: tab-separated")), caption = 'read_delim and derived functions',
  format = "html", table.attr = "style='width:60%;'"
)
```
<br>

Read in a file provided as an example with the `readr` package:

```{r, message=TRUE}
f1 <- read_csv(readr_example("mtcars.csv"))
```

Read in a file available online (https protocol):

```{r, message=TRUE}
f2 <- read_tsv("https://public-docs.crg.es/biocore/projects/training/R_tidyverse_2021/inputB.txt")
```

As you can see, `readr` prints out the **column specifications**, so you can make sure the data is read the way it is meant to be.
<br>

Useful arguments to consider, as you read in the file:
  
* `n_max=k` : read in a subset (first *k* rows).
* col_names:
  * col_names=FALSE : your data doesn't contain headers/column names.
  * col_names=c("A", "B", "C") : you are providing a vector containing column names / header.
* `skip=j` : skip the first *j* rows.

```{r, message=TRUE}
f1 <- read_csv(readr_example("mtcars.csv"),
                n_max=5,
                skip=1,
                col_names=LETTERS[1:11])

f1
```

## Write

`write_delim`, `write_csv`, `write_tsv`: write a `tibble` into a delimited file:

```{r, echo=FALSE, eval=TRUE}
knitr::kable(
  data.frame(name=c("write_delim", "write_csv", "write_tsv"), 
             delim=c("*needs to be set*", "`,`: comma-separated", "`\t`: tab-separated")), caption = 'write_delim and derived functions',
  format = "html", table.attr = "style='width:60%;'"
)
```

```{r, eval=F}
write_csv(x=f2,
          file="test_write_csv.csv")

# to get the exact same in base R, you would need to set a few additional parameters:
write.csv(x=f2,
          file="test_base_write_csv.csv", 
          quote=F, 
          row.names = F)
```

<center><h4 style="background-color: #a4edff; display: inline-block;">**HANDS-ON**</h4></center>

* Read in file `readr_example("mtcars.csv.zip")` (also provided with the readr package)
* With base R coding, remove rows where **mpg > 20**
* Write this new tibble to a **tab-separated file**.

<details>
<summary>
<h5 style="background-color: #a4edff; display: inline-block;">*Answer*</h5>
</summary>

```{r, eval=F}
# Read in file readr_example("mtcars.csv.zip")
Cars <- read_csv(readr_example("mtcars.csv.zip"))

# With base R coding, remove rows where mpg > 20
mpg20 <- Cars[Cars$mpg > 20, ]

# Write this new tibble to a tab-separated file.
write_tsv(mpg20, "mtcars_mpg20.txt")
```

<details>
