# Input / output with *`readr`*

`readr` provides functions for reading in and writing out files.
<br>
Those functions are **up to 10 times faster** that the base R equivalent.

## Read

`read_delim`, `read_csv`, `read_tsv`: Read a delimited file into a `tibble`:

```{r, echo=FALSE, eval=TRUE}
knitr::kable(
  data.frame(name=c("read_delim", "read_csv", "read_tsv"), 
             delim=c("*needs to be set*", "`,`: comma-separated", "`\t`: tab-separated")), caption = 'read_delim and derived functions',
  format = "html", table.attr = "style='width:60%;'"
)
```
<br>

Read in a file provided as an example with the `readr` package:

```{r, message=TRUE}
f1 <- read_csv(readr_example("mtcars.csv"))
```

As you can see, `readr` prints out the column specifications, so you can make sure the data is read the way it is meant to be.
<br>

Useful arguments to consider, as you read in the file:
  
* `n_max` : read in a subset.
* col_names:
  * col_names=FALSE : your data doesn't contain headers/column names.
  * col_names=c("A", "B", "C") : you are providing a vector containing column names / header.
* `skip` : skip rows.

```{r, message=TRUE}
f2 <- read_csv(readr_example("mtcars.csv"),
                n_max=5,
                skip=1,
                col_names=LETTERS[1:11])

f2
```

## Write


format_csv
format_tsv

write_delim
write_csv
write_tsv

**HANDS-ON**

