# Introduction

## What is the "tidyverse"

* A set of packages designed for **data science**:

  * Importing data into R
  * Preparing / Cleaning data
  * Data wrangling (tidying and transforming data)
  * Visualizing data


* All packages share **good practices** in terms of: 
* 
  * Philosophy
  * Grammar
  * Data structure
  
  
## Good reasons to learn (at least a bit) how to use it

* More **intuitive programming**: the names of functions speak for themselves.
* The code is **easier to read** than with R base: it facilitates code sharing.
* More **efficient**: the functions are quite quick (coded more efficiently).
* Very good [**documentation and tutorials**](https://www.tidyverse.org/learn/).

Here is a piece of **R base code**:

```{r class.source="bg-danger", class.output="bg-warning", eval=F}
diamonds2 <- diamonds[diamonds$cut == "Ideal", c("cut", "color", "carat", "price")]
diamonds2[order(diamonds2$price, decreasing=TRUE),]
```

And its equivalent in the **tidyverse**:

```{r df-drop-ok, class.source="bg-success", eval=F}
diamonds %>% 
  select(cut, color, carat, price) %>%
  filter(cut=="Ideal") %>%
  arrange(desc(price))
```


## Tidyverse core packages

The following 8 packages are included in the core **tidyverse version 1.3.0.9000** (current version in April 2021):

* Data Import and Management
  * tibble
  * readr 
  
* Data Wrangling (Tyding and Transformation)
  * dplyr
  * tidyr 
  * stringr
  * forcats


* Data Visualization and Exploration
  * ggplot2


* Functional Programming
  * purrr
  


<br>

```{r, echo=FALSE, eval=TRUE}

tbl_logos <- data.frame(
  logo = "",
  name = c("dplyr", "tidyr", "stringr", "forcats", "readr", "tibble", "purrr", "ggplot2"),
  description=c("Package for data manipulation and exploratory data analysis.", 
                "Package that aims at creating tidy data. Tidy data describe a standard way of storing data.",
                "Package that provides a set of functions for user-friendly string manipulation.",
                "Package that helps you deal with factors", 
                "Package for fast and efficient import and export of data.", 
                "Tibbles are improved - easier to manage -  data frames.", 
                "Package that aims at enhancing R's functional programming toolkit. It provides a set of tools for working with functions and vectors.", 
                "Package for data vizualization of graphics based on Leland Wilkinson's' Grammar of Graphics: graphics are built one layer at a time.")
)

tbl_logos %>%
  kbl(booktabs = T) %>%
  kable_paper(full_width = T) %>%
  column_spec(1, image = spec_image(
    c("images/dplyr_logo.png", "images/tidyr_logo.png", "images/stringr_logo.png", "images/forcats_logo.png", "images/readr_logo.png", "images/tibble_logo.png", "images/purrr_logo.png", "images/ggplot2_logo.png"), width=380, height=270))
    
```


## Install and load all tidyverse package

Install (should be done before the class, as it takes a while!):

```{r, eval=FALSE}
install.packages("tidyverse")
```

Load:

```{r, eval=TRUE}
library(tidyverse)
```

  
  
