# Mini project with code

We will work with two input files:

* **annotation.tsv**: contains the annotation (from Gencode) of 29244 Human genes.
* **normalized_counts.csv**: contains the log2-transformed normalized counts from an RNA-seq project. The expression of 14769 Human genes in **8 samples** from **4 experimental groups** is assessed:
  * Treatment 1, time 0 (2 samples)
  * Treatment 1, time 20 (2 samples)
  * Treatment 2, time 0 (2 samples)
  * Treatment 2, time 20 (2 samples)

1. Download / read in the two files that are found [here](https://public-docs.crg.es/biocore/projects/training/R_tidyverse_2021/mini_project/) into two tibbles.

```{r, eval=FALSE}
# Download and read in the files
annotation <- read_tsv("https://public-docs.crg.es/biocore/projects/training/R_tidyverse_2021/mini_project/annotation.tsv")

norm_data <- read_csv("https://public-docs.crg.es/biocore/projects/training/R_tidyverse_2021/mini_project/normalized_counts.csv")
```

2. **Tidy** each tibble individually.

<details>
<summary>
<h6 style="background-color: #fdffaf; display: inline-block;">*Some tips!*</h6>
</summary>

* For the **annotation**: something needs to be separated.
* For the **normalized data**: it's important to pivot some columns... And to separate one created by the pivoting!

</details>

```{r, eval=FALSE}
# Separate the "gene_name_gene_type" column into gene_name and gene_type
annotation_tidy <- annotation %>% 
                        separate(col=gene_name_gene_type, 
                                 into=c("gene_name", "gene_type"), 
                                 sep="/")

# Pivot the data so there is a single column for the expression value
# Use separate to create 3 new columns that will contain information, respectively on: Treatment, replicate, time point.
# (optional) mutate the "time" column to get a numeric column that contains only the actual time point (without the "t")
norm_data_tidy <- norm_data %>% 
                      pivot_longer(cols=contains("Treatment"), 
                                   values_to="expression", 
                                   names_to="samples") %>% 
                      separate(col=samples, 
                               into=c("Treatment", "replicate", "time"), 
                               sep="_") %>% 
                      mutate(time=as.numeric(str_remove(time, "t")))
```

3. **Join** both datasets so as to obtain one tibble (keep the intersection).

<details>
<summary>
<h6 style="background-color: #fdffaf; display: inline-block;">*Don't know which columns to use for the joining? Click here for help!*</h6>
</summary>

Work on the `gencode_id` column of the normalized data: Gencode IDs only differ from the Ensembl IDs by the suffix (point + numbers).
e.g. ENSG00000140853.15 in Gencode is ENSG00000140853 in Ensembl. Perhaps `str_sub` can help?

</details>

```{r, eval=FALSE}
# First, we need a common ID: remove the suffix with the point + numbers from "gencode_id" in order to obtain "ensembl" IDs
norm_data_tidy <- norm_data_tidy %>% 
                      mutate(ensembl_gene_id=str_sub(gencode_id, 1, 15), 
                             .before=Treatment) # not important, just more convenient

# Use inner join in order to keep only common data
all_data <- inner_join(x=annotation_tidy, 
                       y=norm_data_tidy, 
                       by="ensembl_gene_id")
```

4. What is the **average expression** of the different **types of genes** (`gene_type`)? 
    * According to this data, **which 2 gene types have the highest average expression**? 
    * Remove all rows which correspond to these 2 gene types from the dataset. 
    * What is now the size of our dataset?

<details>
<summary>
<h6 style="background-color: #fdffaf; display: inline-block;">*A couple of tips...*</h6>
</summary>

* Remember `slice_max`? <br>
* `pull` could also be useful!

</details>

```{r, eval=FALSE}
highest_mean_2types <- all_data %>% 
                          group_by(gene_type) %>% 
                          summarise(average_mean_gene_type=mean(expression)) %>% 
                          slice_max(order_by=average_mean_gene_type, 
                                    n=2) %>% 
                          pull(gene_type)

data_no_highest_2types <- all_data %>% 
                            filter(!gene_type %in% highest_mean_2types)

dim(data_no_highest_2types)
```

5. Create a new column that contains **the median expression per gene, per experimental group and per gene type.**<br>
*By experimental group, we mean Treatment + time (for example, samples "Treatment1_rep1_t0" and "Treatment1_rep2_t0" are part of the same experimental group: Treatment1_t0)*

<details>
<summary>
<h6 style="background-color: #fdffaf; display: inline-block;">*Help!*</h6>
</summary>

* If you don't have it already, create a column `experimental_group`.
* The grouping should be done using 3 variables.
* Remember that, if the data is grouped, the newly created columns will take into account the groups...

</details>

```{r, eval=FALSE}
# New column for the actual experimental group (no "rep1" and "rep2")
# group_by gene_name, gene_type and experimental_group
data_no_highest_2_types <- data_no_highest_2types %>% 
                                unite(col="experimental_group", 
                                      Treatment, time, 
                                      sep="_") %>% 
                                group_by(experimental_group, gene_type, gene_name) %>% 
                                mutate(median_expression_experimental_group=median(expression))
```

6. For each experimental group, retrieve the **lincRNA** that has the **highest median expression**.
    * Is it the same lincRNA gene for all 4 experimental groups?

<details>
<summary>
<h6 style="background-color: #fdffaf; display: inline-block;">*Stuck? Click here.*</h6>
</summary>

Check this [stackoverflow post](https://stackoverflow.com/questions/24237399/how-to-select-the-rows-with-maximum-values-in-each-group-with-dplyr) for inspiration.

</details>

```{r, eval=FALSE}
data_no_highest_2_types %>% 
                      ungroup() %>% 
                      group_by(experimental_group, gene_type) %>% 
                      filter(expression == max(expression) & gene_type=="lincRNA")
```


