# Strings manipulation with `stringr`

The `stringr` package provides tools for **string manipulation**.
<br>
All functions in `stringr` start with **str_** and take a vector of strings as the first argument.
<br><br>
We will show here a few useful functions (for a complete list of `stringr` functions, you can have a look at the [Cheat sheet](https://github.com/rstudio/cheatsheets/blob/master/strings.pdf). The cheat sheet also provides guidance on how to work with **regular expressions**.

<br><br>
Let's take a simple **character vector** and a small **tibble** as examples:

```{r, message=FALSE}
examplestring <- c("genomics", "proteomics", "proteome", "transcriptomics", "metagenomics", "metabolomics")

exampletibble <- tibble(day=c("day0", "day1", "day2"),
                    temperature=c("25C", "27C", "24Celsius"))
```

***

* `str_detect`: detects the presence or absence of a pattern in a string.

```{r}
str_detect(examplestring, 
            pattern="genom")
```

You can use regular expressions: as a simple example, here we want to detect which element of `examplestring` **starts** with **genom**.

```{r}
str_detect(examplestring, 
            pattern="^genom")
```

You can reverse the search and output elements where the pattern is NOT found with `negate=TRUE`

```{r}
str_detect(examplestring, 
            pattern="genom",
           negate=TRUE)
```

***

* `str_length`: outputs length of strings (number of characters) in each element of a vector.

```{r}
str_length(examplestring)
```

***

* `str_replace`: looks for a pattern in a string and replace it.

We can replace "omics" with "ome"

```{r}
str_replace(examplestring, 
            pattern="omics", 
            "ome")
```

`str_replace` can be used to remove selected patterns from strings:

```{r}
str_replace(examplestring, 
            pattern="omics", 
            "")

# str_remove is a wrapper for the same thing (no need for the 3rd argument)
str_remove(examplestring, 
            pattern="omics")
```

Same with a **tibble's column**:

```{r}
str_remove(exampletibble$day, 
            pattern="day")
```

You can use it inside another `tidyverse` function:

```{r}
mutate(exampletibble, day=str_remove(day, pattern="day"))
```

***

* `str_count`: count the number of occurences of a pattern:

Count how many times "omics" is found in each element:

```{r}
str_count(examplestring, 
            pattern="omics")
```

Count how many vowels are found in each element:

```{r}
str_count(examplestring, 
            pattern="[aeiouy]")
```

***

* `str_sub`: extracts and replace substrings from a character vector

```{r}
str_sub(examplestring, 
            start=1, # position of the first character
            end=10) # position of the last character
```

Let's keep the first 2 characters of the **temperature** column of our `exampletibble`:

```{r}
str_sub(exampletibble$temperature, 
        start=1, 
        end=2)
```

Within `mutate`:

```{r}
mutate(exampletibble, temperature=str_sub(temperature, start=1, end=2))
```

<center><h4 style="background-color: #a4edff; display: inline-block;">**HANDS-ON**</h4></center>

We will play with the following **character vector**:

```{r}
countries <- c("Germany", "Uganda", "Canada", "Australia", "Switzerland", "Thailand", "Bolivia", "Russia", "Italy", "Senegal", "South Korea", "Mexico", "Argentina", "England")
```

* What is the **average length** of the country names?
* How many country names **end with an "a"**?
* Replace **empty spaces with underscores** in `countries`.
* In which country name do you find more the **letter "a"**?

<details>
<summary>
<h5 style="background-color: #a4edff; display: inline-block;">*Answer*</h5>
</summary>

```{r, eval=F}
# What is the average length of the country names?
mean(str_length(countries))

# How many country names end with an "a"?
  # Get the logical vector
str_detect(countries, "a$")
  # Retrieve only the "TRUE" and count
length(which(str_detect(countries, "a$")))
length(countries[str_detect(countries, "a$")])

# Replace empty spaces with underscores in `countries`.
str_replace(countries, " ", "_")

# In which country name do you find more the letter "a"?
  # count how many "a" per country name
str_count(countries, "a")
  # extract the country name where there are more "a".
countries[max(str_count(countries, "a"))]
```

</details>



