# Tibbles

## What are tibbles?

* Modern re-thinking of data frame.
* They leave behind some features (sometimes user-unfriendly) of classical data frames:
  * No string to factor conversion (*NOTE*: since version 4.0 of R, this is <u>no longer the case for classical data frames</u>)
  * No automatic creation of row names.
  * Column names can be **non-valid R variable names** (for example, they can start with numbers).

You can create a simple tibble as you create a data frame, but with the `tibble` function:

```{r, eval=T}
mytibble <- tibble(
  letters = LETTERS,
  numbers = 1:26
)
```

```{echo=F, eval=F}
mytibble
```

* Convert a data frame to a tibble:

```{r}
mydf <- data.frame(letters = LETTERS,
  numbers = 1:26)
as_tibble(mydf)
```

* Convert a tibble to a data frame (useful to interact with R code that doesn't support tibbles):

```{r}
as.data.frame(mytibble)
```


## Displaying tibbles

The way it **prints** in the console is different that that of data frames:

* Dimensions are shown.
* Information about columns **data types** (no need for `mode()` or `typeof`).
* Only the first 10 rows are displayed.
* Only the columns that **fit the screen** are displayed.

If you want to see more rows, set the `n` parameter in `print()`:

* Print the first 15 rows:

```{r}
print(mytibble, n=15)
```

* Print all rows:

```{r}
print(mytibble, n=Inf)
```

## Subsetting / manipulating tibbles

* Extract column by **name**:

```{r}
# with the $
mytibble$letters

# or the [[]]
mytibble[["letters"]]

# Note: with using simple square brackets, you extract a sub-tibble (2 dimensions)
mytibble["letters"]
```

* Extract column by **position/index**:

```{r}
mytibble
```


**HANDS-ON**

<details>
<summary>
*Answer*
</summary>


</details>




