# Tibbles

## What are tibbles?

* Modern re-thinking of data frame.
* They leave behind some (sometimes user-unfriendly) features of classical data frames:
  * No string to factor conversion (*NOTE*: since version 4.0 of R, this is <u>no longer the case for classical data frames</u>)
  * No automatic creation of row names.
  * Column names can be **non-valid R variable names** (for example, they can start with numbers).

## Creating tibbles

You can create a simple tibble with the `tibble` function, similarly as when you create a data frame:

```{r, echo=TRUE, eval=TRUE}
mytibble <- tibble(
  letters = LETTERS,
  numbers = 1:26
)
```

* Convert a data frame to a tibble:

```{r, eval=F}
mydf <- data.frame(letters = LETTERS,
  numbers = 1:26)
as_tibble(mydf)
```

* Convert a tibble to a data frame (useful to interact with R code that doesn't support tibbles):

```{r, eval=F}
as.data.frame(mytibble)
```


## Displaying tibbles

The way it **prints** in the console is different than that of data frames:

* Dimensions are shown.
* Information about columns **data types** (no need for `mode()` or `typeof`).
* Only the first 10 rows are displayed.
* Only the columns that **fit the screen** are displayed.

If you want to see more rows, set the `n` parameter in `print()`:

* Print the first 15 rows:

```{r}
print(mytibble, n=15)
```

* Print all rows:

```{r}
print(mytibble, n=Inf)
```

## Data types in tibbles

When you display a tibble, note that there is a **3 or 4-letter abbreviation** under each column name that describes the type of each variable:

```{r, echo=F, eval=T}
tbl_dtypes <-tibble(Abbreviation=c("int", "dbl", "chr", "lgl", "fctr", "date", "dttm"), 
                         Description=c("integers", "doubles (or real numbers)", "character vectors (or strings)", "logicals (TRUE or FALSE)", "factors (categorical variables)", "dates", "date-time (a date + a time)"))

tbl_dtypes %>%
  kbl(booktabs = T, align="l") %>%
  kable_paper(full_width = F)
```

## Subsetting / manipulating tibbles

* Extract column by **name**:

```{r}
# with the $
mytibble$letters

# or the [[]]
mytibble[["letters"]]

# Note: with simple square brackets, you extract a sub-tibble (it remains a 2 dimensional object)
mytibble["letters"]
```

* Extract column by **position/index**:

```{r}
# sub-tibble with one column
mytibble[, 1]

# corresponding vector
mytibble[[1]]
```


<center><h4 style="background-color: #a4edff; display: inline-block;">**HANDS-ON**</h4></center>

We will work with the built-in dataset `iris`:

* Is `iris` a tibble? How can you tell? Try to run `is_tibble`.
* Convert `iris` to a tibble. Save in a new object `tbl_iris`. Run `is_tibble` on `tbl_iris`.
* Show all rows of `tbl_iris`.
* Show the first 3 rows of `tbl_iris`.
* Extract column `Species`.

<details>
<summary>
<h5 style="background-color: #a4edff; display: inline-block;">*Answer*</h5>
</summary>

```{r, eval=F}
# Is iris a tibble?
is_tibble(iris)

# Convert iris to a tibble. Save in a new object tbl_iris.
tbl_iris <- tibble(iris)

# Show all rows from tbl_iris.
print(tbl_iris, n=Inf)

# Show the first 3 rows from tbl_iris.
print(tbl_iris, n=3)

# Extract column Species.
  # create a tibble with only one column:
tbl_iris["Species"] 
  # create a vector:
tbl_iris[["Species"]] # same as tbl_iris$Species
```

</details>




